---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Download the data if it does not already exist

````{r}
fileUrl="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#Download the files and unzip it
if (!file.exists("activity.zip")) {
    download.file(fileUrl, method = "curl", destfile = "activity.zip")
}

if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```

Read the data from the **activity.csv** file and convert the dates into R **Date** format

````{r}
#Read the data
activityData <- read.csv("activity.csv",stringsAsFactors=FALSE)
activityData$date <- as.Date(activityData$date,"%Y-%m-%d")
```

Quick summary of the data
```{r} 
summary(activityData)
```

## What is mean total number of steps taken per day?
Make a histogram of the total number of steps taken each day  

```{r}
library(ggplot2)

total.steps<-tapply(activityData$steps, activityData$date, 
                    FUN = sum, na.rm = TRUE)

qplot(total.steps,binwidth = 500,xlab="Total number of steps per day",ylab="Count")
```

Calculate and report the mean and median total number of steps taken per day  

```{r}
library(scales) # for comma function
mean.raw <- mean(total.steps)
median.raw <- median(total.steps)
```
Rounded Mean and Median values  
Mean Steps   = **`r comma(round(mean.raw,0))`**  
Median Steps = **`r comma(median.raw)`**

## What is the average daily activity pattern?

Average number of steps taken across all days  
```{r}
intervalActivity <- tapply(activityData$steps, activityData$interval, mean, na.rm = TRUE)

qplot( as.numeric(names(intervalActivity)),intervalActivity, geom="line",
       xlab="5-Minute interval",ylab="Average across all the days")
```

5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.

```{r}
maxSteps <- max(intervalActivity)
intervalName <- which(intervalActivity==maxSteps)
```

The average max number of steps is **`r maxSteps `** which is for the interval **`r intervalName `** across all days.

## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missingValues<-length(activityData$steps) - sum(complete.cases(activityData))
```

There are **`r missingValues`** missing values in the dataset.

Fill in the missing values using mean for the interval across all days
```{r}
#intervalActivity
newData <- activityData

indexOfNA <- which(is.na(newData$steps))
newData$steps[indexOfNA] <- intervalActivity[as.character(newData$interval[indexOfNA])]


total.steps.new<-tapply(newData$steps, newData$date, FUN = sum)
```

Creat a Histogram for the new dataset (filled in data)
```{r}
qplot(total.steps.new,binwidth = 500,xlab="Total number of steps per day",ylab="Count")
```

Calculate the new mean and medians 
```{r}
mean.new <- mean(total.steps.new)
median.new <- median(total.steps.new)
```

Rounded new Mean and Median values based on imputed data.  
Mean Steps   = **`r comma(round(mean.new,0))`**  
Median Steps = **`r comma(round(median.new,0))`**

Both the **Mean** and **Median** values have gone up in the imputed dataset. 


## Are there differences in activity patterns between weekdays and weekends?

Lets find if a day is a weekday or weekend  
```{r}

newData$day <- format(newData$date,"%a")
newData$weekend <- as.factor(ifelse(newData$day == "Sat" | newData$day == "Sun","Weekend","Weekday"))
```

Lets create a panel plot to compare the two charts (weekday and weekend).   
```{r}
avgSteps <- aggregate(steps ~ interval + weekend, data = newData,FUN=mean)

ggplot(data=avgSteps,aes(y=steps,x=interval,colour=weekend)) + geom_line() + 
    facet_wrap(~weekend,ncol=1) + #facet_grid(weekend~.) +
    xlab("5-minute interval") + ylab("Number of steps") +
    theme(legend.position="none")
```
   
From the chart its clear that there is more walking around on weekends when comared to weekdays.

